Demes: a specification for interchanging complex demography

### 2021-07-29 paper.tex
addressing comments

Here we present ``Demes'', a data model and file format
specification for complex demographic models developed by the PopSim
Consortium. \aprcomment{do we need ``developed by the PopSim Consortium''?}
The Demes data model precisely defines key parameters
of populations and their relationships over time in an extensible way,
and provides a way to explicitly encode all information relevant to
demography while avoiding repetition. This data model is implemented
in the widely used YAML format~\citep{ben2009yaml}, which provides a
good balance between human and machine readability.
The specification precisely
defines the required behaviour of implementations, ensuring that there
is no ambiguity of interpretation, and includes both a reference
implementation and an extensive suite of test examples and their expected
output. The initial software ecosystem includes high-quality
Python and C parser implementations, as well as utilities for verification
and visualisation of Demes models, and has been implemented in
several popular inference and simulation methods
(Table~\ref{tab-software}).
We hope that this
data model and file format will be widely adopted by the community,
such that users can expect to simulate directly from
inferred models with little or no programming effort.

@@

Structurally, the high-level Demes specification encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
Figure~\ref{fig-example} shows an example Demes model expressed in both the
high and low-level forms (both in terms of YAML syntax). For values that repeat
across fields, default may be used to implicity assign a default value to
fields of the given type.  A default is superceded by an explicitly provided
value if given. Other values are inherited naturally following the progression
of time. For example, if an epoch \texttt{start\_size} is not provided, it is
assumed to be equal to the \texttt{end\_size} of the previous epoch. This also
means that the first epoch of each population must specify the initial size.

@@

The Demes specification precisely defines the required behaviour of parsers,
which translate the high-level specifications into unambiguous low-level
representations where all parameters have been assigned values and data
constraints have been checked. The output is formally defined as JSON, but in
practise parsers output an object model that is suitable for the particular
programming language and that can be used directly by the implementing program.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs. In addition, we have high-quality parser implementations in
the Python and C languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

@@

Demes is meant to interface with a large number of simulation and inference
methods, so we have restricted to the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic software, many features of underlying genome biology are
omitted.  Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure). Genome ploidy
is left ambiguous, and selection and dominance models are absent. Using a Demes
model, which specifies deterministic population sizes, as the demographic input
for simulation using software that allows for selection implies that selection
is \emph{soft}, that is, genome fitnesses are rescaled relative to the mean
population fitness \citep{christiansen1975hard}. It is possible to instead
interpret population sizes as carrying capacities, allowing for a model of
\emph{hard} selection. However, this should not be understood as the default
interpretation of population sizes, which conventionally specifies a
deterministic number of individuals.  Finally, as Demes assumes the
exchangeability of individuals within populations, Demes does not provide a way
to define geographic spatial structure within populations.

Demes version 1.0 is intended to provide a stable and minimal specification
that is compatible with a wide range of simulation softwares. Beginning with
version 1.0, a Demes-specified model is expected to provide the version of
Demes for which it is valid, and if unprovided, it should be a valid model
under the default version, 1.0. Future extensions of the Demes specification
will be released with versions following semantic versioning guidelines [cite
or url?], and downstream software that incorporate Demes are expected to
indicate and check Demes version compatibility.

>>>>

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the PopSim Consortium. The Demes data
model precisely defines key parameters of populations and their relationships
over time in an extensible way, and provides a way to explicitly encode all
information relevant to demography while avoiding repetition. This data model
is implemented in the widely used YAML format~\citep{ben2009yaml}, which
provides a good balance between human and machine readability. The
specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output. The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab-software}).  We hope that this
data model and file format will be widely adopted by the community, such that
users can expect to simulate directly from inferred models with little or no
programming effort.

@@

Structurally, the high-level Demes specification encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
Figure~\ref{fig-example} shows an example Demes model expressed in both the
high and low-level forms (both in terms of YAML syntax). For values that repeat
across fields, defaults may be used to implicity assign default values to
fields of the given type.  A default is superceded by an explicitly provided
value if given. Other implicit values are inherited naturally following the
progression of time. For example, if an epoch \texttt{start\_size} is not
provided, it is assumed to be equal to the \texttt{end\_size} of the previous
epoch. This also means that the first epoch of each population must specify the
initial size.

@@

The Demes specification precisely defines the required behaviour of parsers,
which translate the high-level specifications into unambiguous low-level
representations where all parameters have been assigned values and data
constraints have been checked. The output is formally defined as JSON, but in
practice parsers output an object model that is suitable for the particular
programming language and that can be used directly by the implementing program.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs. In addition, we have high-quality parser implementations in
the Python and C languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

@@

Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted to the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic methods, many features of underlying genome biology are
omitted. Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure). Genome ploidy
is left ambiguous, and selection and dominance models are absent. Using a Demes
model, which specifies deterministic population sizes, as the demographic input
for simulation using software that allows for selection implies that selection
is \emph{soft}, that is, genome fitnesses are rescaled relative to the mean
population fitness \citep{christiansen1975hard}. It is possible to instead
interpret population sizes as carrying capacities, allowing for a model of
\emph{hard} selection. However, this should not be understood as the default
interpretation of population sizes, which conventionally specifies a
deterministic number of individuals. Finally, as Demes assumes exchangeability
of individuals within populations, it not provide a way to define geographic
spatial structure within populations.

Demes version 1.0 is intended to provide a stable and minimal specification
that is compatible with a wide range of simulators. To ensure consistency
across implementations, the specification is versioned and we commit to
following semantic versioning guidelines.
